---
title: "Inferring the evolutionary history of the Proboscidea"
author: "Rebecca Cooper"
output: rmarkdown::github_document
vignette: >
  %\VignetteIndexEntry{Inferring the evolutionary history of the     Proboscidea}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, eval = TRUE}
knitr::opts_chunk$set(echo = TRUE, include = TRUE, eval = TRUE)
```
  
# Introduction
    
`DeepDiveR` is an R package for estimating biodiversity through time from fossil occurrence data (Cooper et al 2024). It produces formatted input data and a configuration file that can then be executed in Python through the command line, enabling a clear and reproducible workflow. In this tutorial we will use DeepDiveR to analyse the past diversity of Proboscidea, the order containing elephants and their extinct relatives, using the data of Cantalapiedra et al. (2021). 
  
# Installation  

To install the `DeepDiveR` library, it needs to be downloaded from the GitHub repository using either the package `remotes` or a manual download.
  
```{r install, eval = FALSE}
# Option 1: install the package from GitHub
remotes::install_github("DeepDive-project/DeepDiveR")
# Option 2: load it from a directory after downloading it from
# https://github.com/DeepDive-project/DeepDiveR
# deepdiver_path <- "path_to_DeepDiveR"
# setwd(deepdiver_path)
# library(devtools)
# load_all(".")
```

We can then load the `DeepDiveR` package.

```{r load_packages}
library(DeepDiveR)
```

# Preparing the input data

First we will load our proboscidean occurrence data. This is located in the repository for the workshop, and can be downloaded from the GitHub.

```{r load_data}
# Load proboscidean data
path_dat <- "~/GitHub/cpeg25/deepdive_analysis"
proboscidea <- read.csv(paste0(path_dat, "/Proboscidea_occurrences_DD.csv"))
```

Next, we will view the first few rows of the `data.frame` to get a sense of its contents.

```{r view_data}
head(proboscidea)
```

We can see our 'data.frame' has 2104 occurrences with 5 columns describing the data. 
These columns describing the taxon name, the geographic region where it occurs, the minimum and maximum age range, and a locality identifier. Each row in the `data.frame` corresponds to a single occurrence. This is the standard format of data needed by `DeepDiveR`. The dataset we are using today has already been thoroughly checked through by the authors that compiled it (Cantalapiedra et al. 2021). 

We now need to specify the time bins into which our occurrences will be placed. They do not need to be equally spaced, geological intervals can be used if desired. Here we will use divisions based on the stages of the Cenozoic, split to average to approximately one million year duration each. We will do this by describing a vector which provides the bin boundaries, in millions of years.

```{r set_time}
# Describe vector of bin boundaries
bins <- c(66, 65, 64, 63, 61.6, 60, 59.2, 58.13333, 57.06667, 56, 54.975, 53.95, 52.925, 51.9, 50.875, 
          49.85, 48.825, 47.8, 46.85714, 45.91429, 44.97143, 44.02857, 43.08571, 42.14286, 41.2, 
          40.03667, 38.87333, 37.71, 36.7575, 35.805, 34.8525, 33.9, 32.88667, 31.87333, 30.86, 29.84667, 
          28.83333, 27.82, 26.862, 25.904, 24.946, 23.988, 23.03, 22.16667, 21.30333, 20.44, 19.3225, 
          18.205, 17.0875, 15.97, 14.895, 13.82, 12.725, 11.63, 10.534, 9.438, 8.342, 7.246, 6.2895, 
          5.333, 4.4665, 3.6, 2.58, 0.774, 0)
```

Now we can prepare the input file for `DeepDive`, using the function `prep_dd_input`. Here we need to specify the `data.frame` containing the occurrence data and vector of time bins.
We also want to provide the number of replicates, i.e. how many times we want the occurrences to be placed into the time bins. Here we will specify 10 replicates, to illustrate the process, but a higher number of replicates is needed to capture temporal uncertainty in final analyses e.g. 100 (for a dataset of this size 100 replicates takes approximately 7 minutes to run).
Finally, we also need to give the name of the file that we want to be created. The file takes the form `.csv`, indicating that the values are comma-separated. This will be the file we need to input into `DeepDive`.


```{r create_data, eval = FALSE}
# Create input file for DeepDive
prep_dd_input(
  # Specify occurrence data.frame
  dat = proboscidea,
  # Specify vector containing time bin boundaries
  bins = bins,
  # Specify number of replicates
  r = 10, 
  # Specify name of created file
  output_file = "proboscidea_deepdive_input.csv" 
)
```

# Preparing the configuration file

Now we need to create a configuration file for `DeepDive`, using the function `create_config`. Here we will describe all of the internal settings for the analysis.
Some of the settings must be described by the user. These include...

```{r create_config}
# Create configuration file for DeepDive
config <- create_config(
  # Specify the name for the simulations
  name = "proboscidea",
  # Specify the name of the data file
  data_file = "proboscidea_deepdive_input.csv",
  # Specify vector containing time bin boundaries
  bins = bins,
  # Specify the number of geographic regions to simulate
  n_regions = length(unique(proboscidea$Region))
)
```

Other settings are autofilled by `create_config()`. In order to alter these, we can use the function `edit_config()`. First, we know that there are 3 extant species of elephants (the African savanna elephant _Loxodonta africana_, the African forest elephant _Loxodonta cyclotis_ and the Asian elephant _Elephas maximus_), so let's condition our simulations on this number of living species.

``` {r modify_config}
# Modify the number of extant species ("present_diversity") in simulations
edit_config(config = config,
            module = "general",
            parameter = "present_diversity", 
            value = 3)


edit_config(config = config,
            module = "simulations",
            parameter = "extant_sp", 
            value = c(3, 300))

```

It is possible to allow the geographic regions available in the simulations to change through time. In order to do this, we will create an `regions_matrix` object which will be incorporated in the configuration file. 
Here, we will assume that dispersal of proboscideans to Europe and Asia began between 33.9 and 27 Ma, to North America between 20 and 16 Ma, and to South America between 5.3 and 0.8 Ma. Regions disappearing instead of connecting can be made by setting the label argument to: presence = "FALSE". By default presence = "TRUE" and regions will be connected and remain available until the present day.

```{r geog_limits}
# Create data.frame describing the time for which each region is available
region_ages <- rbind(c("Africa", max(bins), max(bins)), # first proboscideans appear in Africa
                    c("Asia", 33.9, 27),  # Cantalapiedra et al. + Eocene/Oligocene boundary, they can enter Eurasia
                    c("Europe", 33.9, 27),  
                    c("North America", 20, 16),  # Cantalapiedra et al, they can enter North America
                    c("South America", 5.3, 0.8))  # Carrillo et al. and Cantalapiedra et al's data, they can enter South America
region_ages <- as.data.frame(region_ages)

# Label columns
colnames(region_ages) <- c("Region", "MaxAge", "MinAge")

# Connect region data to configuration file
regions_matrix(config = config, region_ages = region_ages)
```

Now we need to describe the technical aspects of the neural network training and analysis process. For a reliable estimate, you will need at least 10,000 training simulations. However, for the purpose of running a quick test analysis, we will use a low number of both training and test simulations. Again we will edit the relevant parameters in the configuration file using the `edit_config` function:

```{r describe_setup}
# Set number of training simulations to 100
edit_config(config = config,
            module = "simulations",
            parameter = "n_training_simulations", 
            value = 1000)

# Set number of test simulations to 10
edit_config(config = config,
            module = "simulations",
            parameter = "n_test_simulations", 
            value = 100)
```

It is possible to train multiple models with varying neural network architectures. Primarily, the number of Long Short-Term Memory layers and nodes can be varied along with the layers and nodes in the dense layer of the network. These can be single numbers e.g. 32, in which case 32 nodes will be implemented in one fully connected layer. Alternatively a vector can be used e.g. c(64, 32). In the latter example, two layers of fully connected nodes would be implemented with 64 nodes and 32 nodes respectively. These settings are added to the configuration file using the function `add_model` which should be called once for each model architecture:

```{r add_model}
# Set number of LSTM and dense nodes for an additional model architecture
add_model(config = config,
          lstm_nodes = c(128, 64),
          dense_nodes = c(128, 64, 32), 
          model_name = "2")

```

Now we are ready to write the configuration file to use in the analysis:

```{r write_config, eval = FALSE}
# Write the configuration file
config$write("proboscidea_config.ini")
```  

# Executing the analysis

Once the configuration and input files are created, the full DeepDive analysis including simulation, model training and predictions can be carried out through a single command line entered in a Terminal (MacOS and Linux) or Command prompt (Windows) window executing the Python script run_dd_config.py.

First we will add some flags to run just a test batch of 100 simulations using -plot_features and -n_sims. These flags cut the run short after simulating and will not produce a trained model. However, running a small test batch of simulations can be a useful first step exercise to check if your parameterization for the simulations is capturing the variation in the distribution of fossil data in the way you would expect. 

Additional flags, -wd and -cpu are available that can allow you to adjust the location of your folder and the number of cpus to run the analysis in the command line. Any settings specified by using flags will overwrite the configuration settings:
  
```{python test run, eval = FALSE, python.reticulate = FALSE}
python .\deepdive-main\run_dd_config.py .\proboscidea_config.ini -plot_features -n_sims 100 -cpu 2
```

You can now navigate to the folder "proboscidea_output_feature_plots" to see plots of the empirical fossil record and of your initial simulations. In an ideal world, the goal is for the empirical fossil record (blue data) to fall within the range of the simulated fossil record (orange). If it does not, you can adjust the parameters of the configuration and repeat this initial simulation step until you are satisfied. Checking that the simulations reflect the distribution of the empirical fossil record helps us to avoid potential sources of extrapolation error.

Now we are ready to run the analysis for real!

  
```{python run, eval = FALSE, python.reticulate = FALSE}
python .\deepdive-main\run_dd_config.py .\proboscidea_config.ini -cpu 2
```

This script will create a "simulations" folder containing the training and test sets of simulations and a "trained_models" folder containing the models and plots of training history. This folder additionally includes plots of features of the empirical and simulated records (e.g. number of occurrences through time and per region, number of localities, fraction of singletons and sampled diversity), CSV files with the predicted diversity trajectories for the test set and for the empirical dataset, and a plot of the estimated diversity trajectory.

# References


Cooper, R. B., Allen, B. J., & Silvestro, D. (2024). DeepDiveR–A software for deep learning estimation of palaeodiversity from fossil occurrences. bioRxiv, 2024-09.


Cantalapiedra, J. L. et al. (2021) The rise and fall of proboscidean ecological diversity. Nat. Ecol. Evol. 5, 1266–1272.
